---
title: "Homework Assignment 1"
author: 
- "Gianluca Fiorini MAT:0001223301"
- "Tommaso Di Bene MAT:0001234555"
- "Nikola Stevanovic MAT:0001231428"
date: "2025-12-02"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
rm(list=ls())
library(knitr)
library(kableExtra)
library(latex2exp)
knitr::opts_chunk$set(echo = TRUE)
```

#### **Acknowledgement**

We accept the code of conduct rules as stated by the professor.

## Data

Let $Y_1,\,...,\,Y_n$ be i.i.d., following a $\text{Uniform}(0,\theta)$
distribution, $\theta>0$. The density common to the random variables is
thus given by: $$f_Y(y;\,\theta)=\begin{cases}
\frac{1}{\theta},\;\; y\in(0;\,\theta) \\
0,\;\; \text{otherwise}
\end{cases}$$ while the cumulative distribution function (CDF) common to
the random variables is given by:$$F_Y(y;\,\theta)=\begin{cases}
0,\;\; y\leq0 \\
\frac{y}{\theta},\;\; 0<y<\theta \\
1,\;\; y\ge\theta
\end{cases}
$$ In the following, we will focus on the maximum estimator for
$\theta$, that is $\hat{\theta}:=Y_{(n)}=\text{max}\{Y_1,\,...,\,Y_n\}$.
In particular, we will use Monte Carlo (MC) simulations to approximate
variance, bias, and MSE of the estimator, and compare those
approximations to exact results derived theoretically.

## Theoretical Results

Since the random variables $Y_1,\,...,\,Y_n$ are i.i.d., with CDF $F_Y$,
the CDF of the maximum estimator $\hat{\theta}$, i.e. the n-th order
statistic for the sample, is given
by:$$F_{Y_{(n)}}(y;\,\theta)=[F_Y(y)]^n=\begin{cases}
0,\;\; y\leq0 \\
(\frac{y}{\theta})^n,\;\; 0<y<\theta \\
1,\;\; y\geq\theta
\end{cases}
$$ From this CDF, we can also recover the density for the maximum
estimator, by simply taking the partial derivative w.r.t. $\text{y}$. So
we get that the density of the maximum estimator $\hat{\theta}$ is given
by:$$
f_{Y(n)}(y;\,\theta)= \frac{\partial}{\partial y}  F_{Y_{(n)}}(y)=\begin{cases}
\frac{n}{\theta^{\,n}}y^{n-1},\;\; y\in(0;\,\theta) \\
0,\;\; \text{otherwise}
\end{cases}$$ We can now use this density to derive the theoretical
quantities of interest, i.e. the bias, the variance, and the MSE of
$\hat{\theta}$.
$$Var(\hat{\theta})=E[(\hat{\theta}-E[\hat{\theta}])^2]=\int_{0}^{\theta}(\hat{\theta}-E[\hat{\theta}])^2\;f_{Y(n)}(y;\,\theta)dy=\frac{n\theta^2}{(n+2)(n+1)^2}$$
$$Bias_\theta(\hat{\theta})=E[\hat{\theta}-\theta]=\int_{0}^{\theta}(\hat{\theta}-\theta)\;f_{Y(n)}(y;\,\theta)dy=-\frac{\theta}{n+1}$$
$$ MSE_\theta(\hat{\theta})=E[(\hat{\theta}-\theta)^2]=Var(\hat{\theta})+Bias_\theta^2(\hat{\theta})=\frac{2\theta^2}{(n+1)(n+2)}$$

## Monte Carlo Simulations

Note that all of the quantities we are interested in can be rewritten in
the form $E[g(\hat{\theta})] = E[g(\max\{Y_1,\,...,\,Y_n\})]$, for some
suitable choice of $\text{g}$. For a chosen value of $\theta$, we can
provide an approximation of $E[g(\hat{\theta})]$ using MC simulations as
follows:

1.  Set some sample size $n$, and a number of samples $B$ to simulate.
2.  For $b=1,\,...,\,B$, generate a sample of size $n$ from the
    $\text{Uniform}(0,\theta)$ distribution.
3.  For each generated sample $Y_{1b},\,...,\,Y_{nb}$, $b=1,\,...,\,B$,
    compute the value of the estimator
    $\hat{\theta}_b:=\text{max}\{Y_{1b},\,...,\,Y_{nb}\}$ and
    $g(\hat{\theta_b})$.
4.  Compute the Monte Carlo estimator for the quantity of interest,
    which is given by:
    $$\hat{E}[g(\hat{\theta})]=\frac{1}{B}\sum_{i=1}^B g(\hat{\theta}_b)$$
    By the law of large numbers,
    $\hat{E}[g(\hat{\theta})] \to E[g(\hat{\theta})] \; \text{as} \; B \to +\infty$

For the quantities of interest, choosing $g$ as appropriate, the Monte
Carlo estimators are given by:
$$ \widehat{Var}(\hat{\theta})=\widehat{E}[(\widehat{\theta}-\widehat{E}[\hat{\theta}])^2]=\frac{1}{B}\sum_{i=1}^B (\hat{\theta}_b- \bar{\hat{\theta}})^2,\;\;\;\;\text{so}\;\;g(\hat{\theta})=(\hat{\theta}- \bar{\hat{\theta}})^2$$

$$\widehat{Bias}_\theta(\hat{\theta})=\widehat{E}[\hat{\theta}-\theta]=\frac{1}{B}\sum_{i=1}^B (\hat{\theta}_b-\theta),\;\;\;\;\text{so}\;\;g(\hat{\theta})=(\hat{\theta}- \theta)$$
$$ \widehat{MSE}_\theta(\hat{\theta})=\widehat{E}[(\hat{\theta}-\theta)^2]=\frac{1}{B}\sum_{i=1}^B (\hat{\theta}_b-\theta)^2,\;\;\;\;\text{so}\;\;g(\hat{\theta})=(\hat{\theta}- {\theta})^2$$
We Now present the code used to perform MC simulations and approximate
the quantities of interest by simulation. In this case we set
$\theta=1$, but any positive value could be used in principle. The
simulation seed was set to "123" for all simulations. We started by
defining a function `MC_sims`, which allows to generate $B$ i.i.d.
samples of size $n$ from any arbitrary distribution and estimate our
quantities of interest, and used it to draw samples from a
$\text{Uniform}(0,\theta)$. Simulations were performed for several
values of $n$, so that we could understand how our estimates are
affected by sample size. For each choice of $n$, we generated $B=1000$
samples, and used them to estimate the quantities of interest at
different sample sizes, as stated above. For each choice of $n$, we also
stored all $B$ realizations of the estimator $\hat{\theta_B}$.

```{r}
#Setting value of theta
theta <- 1

#Setting sample sizes
n <- c(10, 20, 30, 50, 100, 200, 500, 1000)

#Setting number of simulations
B <- 1000
#B <- 10000

###MC simulations
MC_sims <- function(theta, n, B, dist, seed=123) {
  #setting seed
  set.seed(123)
  
  #simulating values from dist
  sims <- dist(n*B, theta) #Generating n*B iid realizations from the desired 
                            #distribution with parameter theta
  
  #Creating a matrix for the results
  sims_mat <- matrix(sims, ncol=n,
                     nrow=B)
  
  #computing the value of the statistic for each sample 
  theta_hat <- apply(sims_mat, 1, max)
  
  #Computing MC estimators for quantities of interest
  #Variance
  var_theta_mc <- var(theta_hat)*(B-1)/B
  
  #Bias
  bias_theta_mc <- mean(theta_hat) - theta
  
  #MSE
  MSE_theta_mc <- mean((theta_hat-theta)^2)
  
  #Reporting results
  return(list(B=B,
              n=n,
              theta=theta,
              theta_sims = theta_hat,
              var_mc = var_theta_mc,
              bias_mc = bias_theta_mc,
              MSE_mc = MSE_theta_mc))

}

#Creating function for sampling from U(0, theta)
dist <- function(size, param) {
  runif(size, min=0, max=param)
}

temp <- MC_sims(theta, n[1], B, dist)

###Storing results for different sample sizes, fixed B and fixed theta

#data frames of different values of theta
theta_df <- data.frame(n=numeric(0), theta_hats=numeric(0))

#data frames of estimated and exact quantities of interest
quant_of_interest <- data.frame(n=numeric(0),
                         var_mc=numeric(0),
                         bias_mc=numeric(0),
                         MSE_mc=numeric(0),
                         var_theo=numeric(0),
                         bias_theo=numeric(0),
                         MSE_theo=numeric(0))


#Generating B iid samples for different choices of n and storing simulated values of estimator and
#quantities of interest
for(i in 1:length(n)){
  temp <- MC_sims(theta, n[i], B, dist)
  theta_df <- rbind(theta_df, data.frame(n=temp$n, theta_hats=temp$theta_sims))
  quant_of_interest <- rbind(quant_of_interest,
                             data.frame(n=temp$n,
                                        var_mc=temp$var_mc,
                                        bias_mc=temp$bias_mc,
                                        MSE_mc=temp$MSE_mc,
                                        var_theo=temp$n*theta^2/((temp$n+2)*(temp$n+1)^2),
                                        bias_theo=-theta/(temp$n+1),
                                        MSE_theo=2*theta^2/((temp$n+1)*(temp$n+2))))
}
```

In the output below we report the MC estimates for the quantities of
interest for different sample sizes (`_mc` variables), accompanied by
the corresponding theoretical values (`_theo` variables).

```{r, echo=F}
  df_tmp1 <- lapply(quant_of_interest[,-1], formatC, format="e", digits=3) |> as.data.frame()
  df_tmp1 <-cbind(n=n, df_tmp1)
  kable(df_tmp1, format = "html", booktabs=T) |> 
  kable_styling("striped", font_size = 12)
```

In the following plot, we present the box plot of the bias distribution
across MC simulations for each sample size. We also include the true,
theoretical values of the finite sample bias. As expected, for smaller
sample sizes the bias distribution is more left skewed, but becomes more
and more concentrated around the true bias, (and in turn towards $0$,
due to the asymptotic behavior of the bias), as $n$ increases.

```{r, fig.dim=c(10,7.5)}
#Finite Sample bias distribution vs theoretical values
theta_df$samp_bias <- theta_df$theta_hats - theta
boxplot(samp_bias ~ n, data=theta_df, pch=19, cex=0.5,
        main="Box plot of Bias Distribution",
        xlab=TeX("$\\hat{theta_{b}}-theta$"),
        ylab=TeX("n"),
        horizontal = T)
points(y=1:length(n),x=quant_of_interest$bias_theo, col = "red", pch=16)
legend(x="topleft", col="red", pch=19, "Theoretical (finite sample) Bias")
```

In the following plots, instead, we provide the MC estimates for the
quantities of interest for the different values of $n$ (black lines),
compared with their true theoretical values (red lines line). In each
plot, for the MC estimates, we also include the simulation error bounds.
For each of the quantities of interest, *approximate* point-wise 95%
simulation error bounds were used, based on the following formula:$$
\text{CI}_{.95}({E}[g(\hat{\theta}))=\widehat{E}[g(\hat{\theta})] \pm z_{\alpha/2}\frac{\hat{\sigma_{\hat{\theta}}}}{\sqrt{B}}$$
Where $\hat{\sigma_{\hat{\theta}}}$ is given by
$\frac{1}{B}\sum_{b=1}^B\bigg(g(\hat{\theta}_b) - \overline{g(\hat{\theta}_b)}\bigg)^2$,
and $g$ was chosen suitably for each quantity, as explained above.

```{r, fig.dim=c(10,6)}
#Computing simulation error bands
alpha <- 0.5
zalpha_2 <- qnorm(alpha/2)

#The simulation error bound
width_bias_mc <- width_MSE_mc <- width_var_mc <- numeric(0)

for (i in 1:length(n)) { #Must be computed separately for different sample sizes
  #For bias
  sigma_hat_bias_mc <- sqrt(var(theta_df$theta_hats[theta_df$n == n[i]] - theta)*(B-1)/B)
  width_bias_mc[i] <- zalpha_2*sigma_hat_bias_mc/sqrt(B) 
  #For MSE
  sigma_hat_MSE_mc <- sqrt(var((theta_df$theta_hats[theta_df$n == n[i]] - theta)^2)*(B-1)/B)
  width_MSE_mc[i] <- zalpha_2*sigma_hat_MSE_mc/sqrt(B) 
  #For Variance
  sigma_hat_var_mc <- sqrt(var((theta_df$theta_hats[theta_df$n == n[i]]-
                                  mean(theta_df$theta_hats[theta_df$n == n[i]]))^2)*(B-1)/B)
  width_var_mc[i] <- zalpha_2*sigma_hat_var_mc/sqrt(B) 
}


###Plotting
#MC Bias vs theoretical bias
plot(1:length(n), quant_of_interest$bias_mc, type="b", lwd = 1.5, xaxt = "n",
     cex=0.8, main="MC Bias vs. Theoretical Bias", ylab="Bias", xlab="n")
lines(1:length(n), quant_of_interest$bias_mc+width_bias_mc, type="l", lwd = 1.5, col = "green")
lines(1:length(n), quant_of_interest$bias_mc-width_bias_mc, type="l", lwd = 1.5, col = "green")
lines(1:length(n), quant_of_interest$bias_theo, type="b", col = "red", lty=2, lwd = 1.5,
      pch=19, cex = 0.4)
axis(1,at=1:length(n),labels=as.character(n))
legend(x="bottomright", c( "MC estimate of Bias","95% Simulation error bounds", "Theoretical Bias"), lty=c(1,1,2), col=c("black", "green", "red"))

#MC Var vs theoretical Var
plot(1:length(n), quant_of_interest$var_mc, type="b", lwd = 1.5,  xaxt = "n",
      cex=0.8, main="MC Variance vs. Theoretical Variance", ylab="Variance", xlab="n")
lines(1:length(n), quant_of_interest$var_mc+width_var_mc, type="l", lwd = 1.5, col = "green")
lines(1:length(n), quant_of_interest$var_mc-width_var_mc, type="l", lwd = 1.5, col = "green")
lines(1:length(n), quant_of_interest$var_theo, type="b", col = "red", lty=2, lwd = 1.5,
      pch=19, cex = 0.4)
axis(1,at=1:length(n),labels=as.character(n))
legend(x="topright", c( "MC estimate of Bias","95% Simulation error bounds", "Theoretical Bias"), lty=c(1,1,2), col=c("black", "green", "red"))

#MC MSE vs theoretical MSE
plot(1:length(n), quant_of_interest$MSE_mc, type="b", lwd = 1.5,  xaxt = "n",
     cex=0.8, main="MC MSE vs. Theoretical MSE", ylab="MSE", xlab="n")
lines(1:length(n), quant_of_interest$MSE_mc+width_MSE_mc, type="l", lwd = 1.5, col = "green")
lines(1:length(n), quant_of_interest$MSE_mc-width_MSE_mc, type="l", lwd = 1.5, col = "green")
lines(1:length(n), quant_of_interest$MSE_theo, type="b", col = "red", lty=2, lwd = 1.5,
      pch=19, cex = 0.4)
axis(1,at=1:length(n),labels=as.character(n))
legend(x="topright", c( "MC estimate of MSE","95% Simulation error bounds", "Theoretical MSE"), lty=c(1,1,2), col=c("black", "green", "red"))



```

## Additional Questions

1.  To run MC simulations we needed to make choices concerning 3
    parameters: $\theta$, $n$, and $B$. $\theta$ is the population
    parameter that governs our distribution. It is assumed known in MC
    simulations, so we arbitrarily set $\theta=1$, although any positive
    value could have been used. (To answer one of the following
    questions we did in fact perform simulations using different choices
    of $\theta$). $B$ is the number of samples we simulate from the
    population. In order to get reliable MC estimates of the quantities
    of interest it must be fairly large, so we chose $B=1000$. Of
    course, larger values would've led to more accurate MC estimates.
    Finally, $n$ is the sample size, i.e. the size of each of the $B$
    samples that we generate in MC simulations. We chose several values
    of $n$ to perform simulations so that we could understand the
    behavior of the MC estimator as the sample size varies, while $B$
    and $\theta$ are kept fixed.

    An additional parameter that could be set is the simulation seed,
    which is needed for reproducibility. We set the simulation seed to
    $123$, as requested by the assignment.

2.  Yes, there are differences in the accuracy of MC simulations across
    different sample sizes. This can clearly be seen from the plots
    presented above, where, as the sample size increases, the MC
    estimates get closer and closer to the theoretical quantities, and
    the simulation error bounds become tighter and tighter. This
    intuitively makes sense, because, as $B$ and $\theta$ are kept
    fixed, increases in the sample size mean that, for each of the $B$
    simulated samples, the simulated $\hat{\theta_b}$ will be closer to
    the true value of the parameter on average, and present a lower
    amount of variability. This means that our MC estimates, although
    based on the same number of simulated values, will actually involve
    more accurate estimates $\hat{\theta_b}$ for large $n$, and in turn
    provide better accuracy themselves.

3.  To answer this question, we fixed $B=1000$ and $n=100$, then ran MC
    simulations for several choices of $\theta$, and produced plots
    analogous for the ones used above to gauge the effect of different
    parameter values on the MC estimates of our quantities of interest.
    Other values of $n$ were also tried [not reported].

    As we can see from these plots, it seems that for fixed $n$ and $B$,
    the MC estimates become less accurate as $\theta$ increases. The
    loss in accuracy as $\theta$ grows becomes more pronounced for
    smaller values of $n$. This makes sense because, by increasing the
    value of $\theta$, we are widening the support of the underlying
    distribution ($\text{Uniform}(0,\theta)$) and introducing additional
    variability, which in turn reduces the accuracy of our MC
    simulations. By widening the support of the distribution while
    keeping $n$ and $B$ fixed we lose accuracy because it becomes more
    difficult to adequately explore the new support with the same number
    of simulated values.

```{r, fig.dim=c(10,7)}
###MC Simulations for different theta values
theta <- c(0.1,0.25,0.5,1,1.5,2,5,10)
n <- 100
B <- 1000

dist <- function(size, param) {
  runif(size, min=0, max=param)
}

#data frames of different values of theta
theta_df <- data.frame(n=numeric(0), theta=numeric(0), theta_hats=numeric(0))

#data frames of estimated and exact quantities of interest
quant_of_interest <- data.frame(n=numeric(0),
                                var_mc=numeric(0),
                                bias_mc=numeric(0),
                                MSE_mc=numeric(0),
                                var_theo=numeric(0),
                                bias_theo=numeric(0),
                                MSE_theo=numeric(0))

for(i in 1:length(theta)){
  temp <- MC_sims(theta[i], n, B, dist)
  theta_df <- rbind(theta_df, data.frame(n=temp$n, theta=temp$theta, theta_hats=temp$theta_sims))
  quant_of_interest <- rbind(quant_of_interest,
                             data.frame(n=temp$n,
                                        theta=temp$theta,
                                        var_mc=temp$var_mc,
                                        bias_mc=temp$bias_mc,
                                        MSE_mc=temp$MSE_mc,
                                        var_theo=temp$n*temp$theta^2/((temp$n+2)*(temp$n+1)^2),
                                        bias_theo=-temp$theta/(temp$n+1),
                                        MSE_theo=2*temp$theta^2/((temp$n+1)*(temp$n+2))))
}

#Finite Sample bias distribution vs theoretical values
theta_df$samp_bias <- theta_df$theta_hats - theta_df$theta
boxplot(samp_bias ~ theta, data=theta_df, pch=19, cex=0.5, alpha = 0.5,
        main="Box plot of Bias Distribution",
        xlab=TeX("$\\hat{theta_{b}}-theta$"),
        ylab=TeX("$\\theta"),
        horizontal = T)
points(y=1:length(theta),x=quant_of_interest$bias_theo, col = "red", pch=16)
legend(x="bottomleft", col="red", pch=19, "Theoretical (finite sample) Bias")

#Computing simulation error bands
alpha <- 0.5
zalpha_2 <- qnorm(alpha/2)

w_theta_MSE <- w_theta_bias <- w_theta_var <- numeric(0)

#The simulation error bound
for (i in 1:length(theta)) {
  #For bias
  sigma_hat_bias_mc <- sqrt(var(theta_df$theta_hats[theta_df$theta == theta[i]] - theta[i])*(B-1)/B)
  w_theta_bias[i] <- zalpha_2*sigma_hat_bias_mc/sqrt(B) 
  #For MSE
  sigma_hat_MSE_mc <- sqrt(var((theta_df$theta_hats[theta_df$theta == theta[i]] - theta[i])^2)*(B-1)/B)
  w_theta_MSE[i] <- zalpha_2*sigma_hat_MSE_mc/sqrt(B) 
  #For Variance
  sigma_hat_var_mc <- sqrt(var((theta_df$theta_hats[theta_df$theta == theta[i]]-
                                  mean(theta_df$theta_hats[theta_df$theta == theta[i]]))^2)*(B-1)/B)
  w_theta_var[i] <- zalpha_2*sigma_hat_var_mc/sqrt(B) 
}


#MC Bias vs theoretical bias
plot(1:length(theta), quant_of_interest$bias_mc, type="b", lwd = 1.5, xaxt = "n",
     cex=0.8, main="MC Bias vs. Theoretical Bias", ylab="Bias", xlab=TeX("$\\theta"))
lines(1:length(theta), quant_of_interest$bias_mc+w_theta_bias, type="l", lwd = 1.5, col = "green")
lines(1:length(theta), quant_of_interest$bias_mc-w_theta_bias, type="l", lwd = 1.5, col = "green")
lines(1:length(theta), quant_of_interest$bias_theo, type="b", col = "red", lty=2, lwd = 1.5,
      pch=19, cex = 0.4)
axis(1,at=1:length(theta),labels=as.character(theta))
legend(x="bottomleft", c( "MC estimate of Bias","95% Simulation error bounds", "Theoretical Bias"), lty=c(1,1,2), col=c("black", "green", "red"))

#MC Var vs theoretical Var
plot(1:length(theta), quant_of_interest$var_mc, type="b", lwd = 1.5, xaxt = "n",
     cex=0.8, main="MC Variance vs. Theoretical Variance", ylab="Variance", xlab=TeX("$\\theta"))
lines(1:length(theta), quant_of_interest$var_mc+w_theta_var, type="l", lwd = 1.5, col = "green")
lines(1:length(theta), quant_of_interest$var_mc-w_theta_var, type="l", lwd = 1.5, col = "green")
lines(1:length(theta), quant_of_interest$var_theo, type="b", col = "red", lty=2, lwd = 1.5,
      pch=19, cex = 0.4)
axis(1,at=1:length(theta),labels=as.character(theta))
legend(x="topleft", c( "MC estimate of Variance","95% Simulation error bounds", "Theoretical Variance"), lty=c(1,1,2), col=c("black", "green", "red"))

#MC MSE vs theoretical MSE
plot(1:length(theta), quant_of_interest$MSE_mc, type="b", lwd = 1.5, xaxt = "n",
     cex=0.8, main="MC MSE vs. Theoretical MSE", ylab="MSE", xlab=TeX("$\\theta"))
lines(1:length(theta), quant_of_interest$MSE_mc+w_theta_MSE, type="l", lwd = 1.5, col = "green")
lines(1:length(theta), quant_of_interest$MSE_mc-w_theta_MSE, type="l", lwd = 1.5, col = "green")
lines(1:length(theta), quant_of_interest$MSE_theo, type="b", col = "red", lty=2, lwd = 1.5,
      pch=19, cex = 0.4)
axis(1,at=1:length(theta),labels=as.character(theta))
legend(x="topleft", c( "MC estimate of MSE","95% Simulation error bounds", "Theoretical MSE"), lty=c(1,1,2), col=c("black", "green", "red"))


```

4.  Based on the plots we computed earlier, the difference between the
    MC estimates of bias, variance, and MSE, accounting for the
    simulation error, and the true theoretical values is large for small
    $n$, and decreases as the sample size grows, (assuming $B$ and
    $\theta$ fixed). This is expected, because, as mentioned earlier, by
    increasing $n$ we are working with higher-quality, (i.e. more
    accurate and less variable), estimates of the parameter from each
    simulated sample, so that the convergence of the MC estimator
    towards the true theoretical value occurs more quickly as
    $B \to +\infty$. Clearly, this implies that while keeping $B$ fixed
    and increasing $n$ we are still improving the quality of our MC
    estimates. Moreover, for analogous reasons, increases in the value
    of $n$ while $B$ remains fixed also reduce the simulation error
    bounds, which means that we are more and more confident in our
    estimates.
